---
title: "UW Summer Institutes - Introduction to Metagenomic Data Analysis - Lab 2"
output: html_notebook
---

# Getting ready

```{r load_packages, message=FALSE}
library(dada2); packageVersion("dada2")
library(phyloseq); packageVersion("phyloseq")

library(stringr); packageVersion("stringr")
library(ggplot2); packageVersion("ggplot2")
library(dplyr); packageVersion("dplyr")
```

## Load data from Nature 488, pp. 621-626: `STAT.RData`

```{r load_data}
path <- "../data/STAT.rdata"
load(path)
phy
head(phenotypes)
tail(phenotypes)
```

```{r inspect_sample_data}
head(sample_data(phy))
levels(sample_data(phy)$Treatment)
```

## Merge `phenotypes` variable with the	`phyloseq` data

```{r pheno}
unique(sample_names(phy))

pheno1 <- phenotypes
pheno2 <- phenotypes

rownames(pheno1) <- paste('cecal', rownames(pheno1), sep = "_")
rownames(pheno2) <- paste('fecal', rownames(pheno2), sep = "_")
pheno <- rbind(pheno1, pheno2)

head(pheno)
tail(pheno)
# phenotypes_fixed <- phenotypes
# rownames(phenotypes_fixed) <- str_replace_all(rownames(phenotypes), 
#                                               "C", "cecal_C")
# rownames(phenotypes_fixed) <- str_replace_all(rownames(phenotypes_fixed), 
#                                               "C", "cecal_C")
```

```{r phy2}
phy2 <- merge_phyloseq(phy, sample_data(pheno))
phy
phy2
head(sample_data(phy2))
```

## Estimate observed species with Chao1 and Shannon diversity using `estimate_richness()` function

```{r alpha_div}
alpha_meas <- c("Observed", "Chao1", "Shannon")
alpha_div <- estimate_richness(phy2, measures = alpha_meas)
glimpse(alpha_div)
```

## Subsetting

```{r phy3}
phy3 <- subset_samples(phy2, sample_sums(phy2) > 3500)
phy3
```

```{r alpha_div_sub}
alpha_div_sub <- estimate_richness(phy3, measures = alpha_meas)
```


## Estimate rarified diversities in #3 averaged over 20 replicates and compare the two estimates

```{r alpha_div_rare}
set.seed(0)
alpha_div_rare <- matrix(0, ncol = ncol(alpha_div), nrow = nrow(alpha_div))
for (i in 1:20) {
    alpha_div_rare <- alpha_div_rare +
        estimate_richness(
            rarefy_even_depth(phy3, sample.size = 3500, 
                              verbose = F, trimOTUs = F),
            measures = alpha_meas
        )
}
alpha_div_rare <- alpha_div_rare / 20
glimpse(alpha_div_rare)
```

```{r plot_compare}
plot(alpha_div$Observed, alpha_div_rare$Observed)
plot(alpha_div$Chao1, alpha_div_rare$Chao1)
plot(alpha_div$Shannon, alpha_div_rare$Shannon)
```


## Summarize the data at the order level

```{r order_phy}
order_phy <- tax_glom(phy3, taxrank = "Order")
glimpse(order_phy)
```

## Normalize the data using CLR, relative abundance, and DESeq2

```{r order_rel}
order_rel <- transform_sample_counts(order_phy,
                                     function(x) x / sum(x))
```

```{r order_clr}
order_clr <- transform_sample_counts(order_phy,
                                     function(x) {
                                         y = log(x + 1)
                                         y / sum(y)
                                     })
```

```{r order_deseq}
library(DESeq2)
countData = round(as(otu_table(order_phy), "matrix"), digits = 0)
countData = countData + 1L
dds <- DESeqDataSetFromMatrix(countData, sample_data(order_phy), design = ~1)

order_deseq = merge_phyloseq(otu_table(counts(estimateSizeFactors(dds), 
                                              normalized=T), taxa_are_rows = T), 
                             sample_data(order_phy), 
                             phy_tree(order_phy), 
                             tax_table(order_phy))
glimpse(order_deseq)
```

## Compute appropriate univariate tests on normalized data

```{r univ.location.res}
Location = sample_data(order_rel)$Location
rel.p = apply(otu_table(order_rel),1, function(x) wilcox.test(c(x)~Location)$p.value)
clr.p = apply(otu_table(order_clr),1, function(x) t.test(c(x)~Location)$p.value)
deseq.p = apply(otu_table(order_deseq),1, function(x) t.test(log10(c(x))~Location)$p.value)

univ.location.res = data.frame(taxa = tax_table(order_clr)[,"Order"], rel.p, clr.p, deseq.p)
head(univ.location.res)
```

## Adjust the p-values using False Discovery Rate

```{r do_fdr}
univ.location.res$rel.fdr = p.adjust(univ.location.res$rel.p, method="fdr")
univ.location.res$clr.fdr = p.adjust(univ.location.res$clr.p, method="fdr")
univ.location.res$deseq.fdr = p.adjust(univ.location.res$deseq.p, method="fdr")

colSums(univ.location.res<0.05)
univ.location.res
```

## Plot the abundances using `plot_heatmap()`, `plot_bar()`, or produce box and whisker plots

```{r plot_heatmap_rel}
plot_heatmap(order_rel, taxa.label = "Order", sample.order = "Location")+
  facet_grid(.~Location, scales = "free_x")
```

```{r plot_heatmap_clr}
plot_heatmap(order_clr, taxa.label = "Order", sample.order = "Location")+
  facet_grid(.~Location, scales = "free_x")
```

```{r plot_heatmap_deseq}
plot_heatmap(order_deseq, taxa.label = "Order", sample.order = "Location")+
  facet_grid(.~Location, scales = "free_x")
```

```{r plot_barplot_rel, fig.height=6}
plot_bar(order_rel, fill="Location") + 
  facet_wrap(facets = "Order", ncol=5, nrow=5, scales = "free_y")
```

```{r plot_barplot_clr, fig.height=6}
plot_bar(order_clr, fill="Location") + 
  facet_wrap(facets = "Order", ncol=5, nrow=5, scales = "free_y")
```

```{r plot_barplot_deseq, fig.height=6}
plot_bar(order_deseq, fill="Location") + 
  facet_wrap(facets = "Order", ncol=5, nrow=5, scales = "free_y")
```

```{r plot_boxplot_rel, fig.height=6}
ggplot(psmelt(order_rel), aes(x=Location, y=Abundance)) + 
  geom_boxplot(aes(fill=Location)) + geom_jitter() + 
  facet_wrap(facets="Order", ncol=5, nrow=5, scales="free_y")
```

```{r plot_boxplot_clr, fig.height=6}
ggplot(psmelt(order_clr), aes(x=Location, y=Abundance)) + 
  geom_boxplot(aes(fill=Location)) + geom_jitter() + 
  facet_wrap(facets="Order", ncol=5, nrow=5, scales="free_y")
```

```{r plot_boxplot_deseq, fig.height=6}
ggplot(psmelt(order_deseq), aes(x=Location, y=Abundance)) + 
  geom_boxplot(aes(fill=Location)) + geom_jitter() + 
  facet_wrap(facets="Order", ncol=5, nrow=5, scales="free_y")
```

## Subset the order level data to only fecal samples

```{r fecal.rel}
fecal.rel = subset_samples(order_rel, Location = "fecal")
```

## Filter out taxa with low abundance (mean abundance < 0.1%)

```{r fecal.fel.subs}
fecal.rel.subs = subset_taxa(fecal.rel, rowMeans(otu_table(fecal.rel))> 0.001)
```

## Perform univariate analysis of the fecal subset with respect to the Treatment variable





