---
title: "Lab 02: Quality Control, Preprocessing, and Univariate Testing -- Example Solutions"
author: SISMID 2016 Module 14
date: "`r date()`"
output: 
  html_document: 
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: true
---


--------------

# Load the data

> Load data from Nature 488, pp. 621-626: `STAT.RData`

```{r}
load('STAT.RData')
library(phyloseq)
```

```{r}
head(sample_data(phy))
levels(sample_data(phy)$Treatment)
```

```{r}
head(phenotypes)
phy
sample_names(phy)[1:5]
```

The sample names are in different formats: phenotypes are per mouse; phy data are per mouse-location.

&nbsp;

---

# Merge phenotypes

> Merge `phenotypes` variable with the	`phyloseq` data

```{r}
pheno1 = phenotypes
pheno2 = phenotypes

rownames(pheno1) = paste('cecal', rownames(pheno1), sep="_")
rownames(pheno2) = paste('fecal', rownames(pheno2), sep="_")
pheno = rbind(pheno1, pheno2)

phy2 = merge_phyloseq(phy, sample_data(pheno))

phy
phy2
head(sample_data(phy2))
```

```{r, echo=FALSE}
## cleanup
rm(pheno, pheno1, pheno2, phy, phenotypes)
```

&nbsp;

---

# Estimate richness

> Estimate observed species with Chao1 and Shannon diversity using `estimate_richness()` function

```{r, eval=FALSE}
?estimate_richness
```

```{r}
alpha = estimate_richness(phy2, measures=c("Observed", "Chao1", "Shannon"))
head(alpha)
```

```{r}
plot(ecdf(sample_sums(phy2)))
```

&nbsp;

---

# Rarefied diversities

> Estimate rarified diversities in #3 averaged over 20 replicates and compare the two estimates

## Let's only keep samples that have at least 3500 observations/reads

```{r}
phy3 = subset_samples(phy2, sample_sums(phy2)>3500)
alpha = estimate_richness(phy3, measures=c("Observed", "Chao1", "Shannon"))

```

## Estimate rarefied diversities

```{r}
alpha.rare = matrix(0, ncol=ncol(alpha), nrow=nrow(alpha))
for(i in 1:20){
  alpha.rare= alpha.rare + 
    estimate_richness(rarefy_even_depth(phy3, sample.size = 3500,
                                        verbose=F, trimOTUs = F), 
                      measures=c("Observed", "Chao1", "Shannon"))
}
alpha.rare = alpha.rare/20
```

## Compare the rarefied vs non-rarefied alpha diversity

```{r}
plot(alpha$Observed, alpha.rare$Observed)
plot(alpha$Chao1, alpha.rare$Chao1)
plot(alpha$Shannon, alpha.rare$Shannon)
```

&nbsp;

---

# Order-level abundances

## Compute order level abundances

> Summarize the data at the order level

```{r}
order.phy = tax_glom(phy3, taxrank = "Order")
```

## Normalize the data

> Normalize the data using CLR, relative abundance, and DESeq2

### relative abundance

```{r}
order.rel = transform_sample_counts(order.phy, function(x) x/sum(x))
```

### clr

```{r}
order.clr = transform_sample_counts(order.phy, function(x){y=log(x+1); y/sum(y)})
```

### DESeq2

```{r}
library(DESeq2)
countData = round(as(otu_table(order.phy), "matrix"), digits = 0)
countData = countData + 1L
dds <- DESeqDataSetFromMatrix(countData, sample_data(order.phy), design = ~1)

order.deseq = merge_phyloseq(otu_table(counts(estimateSizeFactors(dds), 
                                              normalized=T), taxa_are_rows = T), 
                             sample_data(order.phy), 
                             phy_tree(order.phy), 
                             tax_table(order.phy))
```


```{r, echo=FALSE}
## cleanup
rm(dds, countData)
```

&nbsp;

---

# Univariate tests

> Compute appropriate univariate tests on normalized data

## Compare location

```{r}
Location = sample_data(order.rel)$Location
rel.p = apply(otu_table(order.rel),1, function(x) wilcox.test(c(x)~Location)$p.value)
clr.p = apply(otu_table(order.clr),1, function(x) t.test(c(x)~Location)$p.value)
deseq.p = apply(otu_table(order.deseq),1, function(x) t.test(log10(c(x))~Location)$p.value)

univ.location.res = data.frame(taxa = tax_table(order.clr)[,"Order"], rel.p, clr.p, deseq.p)
```

## do FDR

> Adjust the p-values using False Discovery Rate

```{r}
univ.location.res$rel.fdr = p.adjust(univ.location.res$rel.p, method="fdr")
univ.location.res$clr.fdr = p.adjust(univ.location.res$clr.p, method="fdr")
univ.location.res$deseq.fdr = p.adjust(univ.location.res$deseq.p, method="fdr")

colSums(univ.location.res<0.05)
univ.location.res
```


## Plot abundances

> Plot the abundances using `plot_heatmap()`, `plot_bar()`, or produce box and whisker plots

### Plot heatmaps

```{r}
library(ggplot2)
plot_heatmap(order.rel, taxa.label = "Order", sample.order = "Location")+
  facet_grid(.~Location, scales = "free_x")
plot_heatmap(order.clr, taxa.label = "Order", sample.order = "Location")+
  facet_grid(.~Location, scales = "free_x")
plot_heatmap(order.deseq, taxa.label = "Order", sample.order = "Location")+
  facet_grid(.~Location, scales = "free_x")
```


### Plot barplots of all data

```{r}
plot_bar(order.rel, fill="Location") + 
  facet_wrap(facets = "Order", ncol=7, nrow=3, scales = "free_y")
plot_bar(order.clr, fill="Location") + 
  facet_wrap(facets = "Order", ncol=7, nrow=3, scales = "free_y")
plot_bar(order.deseq, fill="Location") + 
  facet_wrap(facets = "Order", ncol=7, nrow=3, scales = "free_y")
```

### Plot box and whiskers charts

```{r}
ggplot(psmelt(order.rel), aes(x=Location, y=Abundance)) + 
  geom_boxplot(aes(fill=Location)) + geom_jitter() + 
  facet_wrap(facets="Order", ncol=7, nrow=3, scales="free_y")

ggplot(psmelt(order.clr), aes(x=Location, y=Abundance)) + 
  geom_boxplot(aes(fill=Location)) + geom_jitter() + 
  facet_wrap(facets="Order", ncol=7, nrow=3, scales="free_y")

ggplot(psmelt(order.deseq), aes(x=Location, y=Abundance)) + 
  geom_boxplot(aes(fill=Location)) + geom_jitter() + 
  facet_wrap(facets="Order", ncol=7, nrow=3, scales="free_y") + scale_y_log10()
```


&nbsp;

---

# Testing in fecal samples

## Subset to only fecal samples

> Subset the order level data to only fecal samples

```{r}
fecal.rel = subset_samples(order.rel, Location = "fecal")
```

## Filter low abundance taxa

> Filter out taxa with low abundance (mean abundance < 0.1%)

```{r}
fecal.rel.subs = subset_taxa(fecal.rel, rowMeans(otu_table(fecal.rel))> 0.001)
```

## Fecal samples vs Treatment

> Perform univariate analysis of the fecal subset with respect to the Treatment variable

```{r}
Treatment = sample_data(fecal.rel.subs)$Treatment

fecal.res = data.frame(taxa = tax_table(fecal.rel.subs)[,"Order"], 
                               fecal.p = apply(otu_table(fecal.rel.subs),1, 
                                               function(x) kruskal.test(c(x)~Treatment)$p.value))
fecal.res$fecal.fdr = p.adjust(fecal.res$fecal.p, method="fdr")

ggplot(psmelt(fecal.rel.subs), aes(x=Treatment, y=Abundance)) + 
  geom_boxplot(aes(fill=Treatment)) + geom_jitter() + 
  facet_wrap(facets="Order", ncol=4, nrow=2, scales="free_y")
```

## Fecal samples vs BMD

```{r}
fecal.deseq = subset_samples(order.deseq, Location = "fecal")

pheno.res = data.frame(taxa = tax_table(fecal.deseq)[,"Order"], 
                       BMD.p = apply(otu_table(fecal.deseq),1, 
                                       function(x) cor.test(log(c(x)), sample_data(fecal.deseq)$BMD)$p.value))
pheno.res$BMD.fdr = p.adjust(pheno.res$BMD.p, method="fdr")

pheno.res$pFAT.p = apply(otu_table(fecal.deseq),1, 
                         function(x) cor.test(log(c(x)), sample_data(fecal.deseq)$pFat)$p.value)

pheno.res$pFAT.fdr = p.adjust(pheno.res$pFAT.p, method="fdr")


```

### Compute global FDR

```{r}
gFDR = with(pheno.res, 
            matrix(p.adjust(c(BMD.p, pFAT.p), method="fdr"), 
                   ncol=2, nrow=nrow(pheno.res), byrow=F))
gFDR
```


### Plot phenotypic variables vs. "the most significant" result

```{r}
ggplot(subset(psmelt(fecal.deseq), Order=="Enterobacteriales"), aes(x=BMD, y=Abundance)) +
  geom_point() + scale_y_log10()
ggplot(subset(psmelt(fecal.deseq), Order=="Enterobacteriales"), aes(x=pFat, y=Abundance)) + 
  geom_point() + scale_y_log10()

```


### Plot phenotypic variables vs. Treatment

```{r}
ggplot(sample_data(fecal.deseq), aes(x=Treatment, y=BMD)) + 
  geom_boxplot(aes(fill=Treatment))+geom_jitter()
ggplot(sample_data(fecal.deseq), aes(x=Treatment, y=pFat)) + 
  geom_boxplot(aes(fill=Treatment))+geom_jitter()

```




