---
title: "Lab 03: Shot Noise Poisson Simulation (for intuition)"
author: SISMID 2016 Module 14
date: "`r date()`"
output: 
  html_document: 
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: true
---


&nbsp;

---

# Load packages

```{r loadpackages}
library("ggplot2"); packageVersion("ggplot2")
theme_set(theme_bw())
library("plyr"); packageVersion("plyr")
library("data.table"); packageVersion("data.table")
set.seed(711L)
```


&nbsp;

---

# Define simple microbiome

Let's define an imaginary and simple vector of proportions of bacterial taxa
in a community.

```{r}
OTUproportions = c(0.3, 0.2, 0.1,
                   rep(0.05, 5), 
                   rep(0.025, 6), 
                   rep(0.001, 6), 
                   rep(0.0001, 6))
names(OTUproportions) <- LETTERS[1:length(OTUproportions)]
OTUproportions
sum(OTUproportions)
# Plot it to be clear.
spdf = data.frame(proportion=OTUproportions, Taxa=names(OTUproportions))
p = ggplot(spdf, aes(Taxa, proportion)) + geom_bar(stat="identity") 
p + ggtitle("The 'true' Taxa proportions")
```

Now sample from these true proportions, and plot the uncertainty. This uses the `sample` function in R.

```{r simulate}
sim1 = sample(x=names(OTUproportions), size=100, replace=TRUE, prob=OTUproportions)
sim1
sim1df = data.frame(table(sim1))
sim1df
sim1df$proportion <- sim1df$Freq/sum(sim1df$Freq)
colnames(sim1df)[1] <- "Taxa"
# Add simulated proportions to the plot of the true proportions.
title = "One realization of multinomial sampling process"
p + geom_point(data=sim1df, color="blue", size=6) +
  ggtitle(title)
```


&nbsp;

---

# Define simulation

Define and test a function that simulates from this

```{r}
simulate_community = function(props, size){
  # props = OTUproportions
  # size = 100
  simi = sample(x=names(props), size=size, replace=TRUE, prob=props)
  simidf = data.table(table(simi))
  simidf[, proportion := N / sum(N, na.rm = TRUE)]
  setnames(simidf, "simi", "Taxa")
  return(simidf)
}
# Test the function
show(simulate_community(OTUproportions, 200))
# Repeat twice, store as list
replicate(2, simulate_community(OTUproportions, 100), simplify = FALSE)
```


&nbsp;

---

# Repeat Simulation

Now repeat the simulated Taxa sampling 
many times at different *library sizes*.

```{r}
N = 300
LibrarySizes = c(100, 2000, 30000, 400000)
bigdt = rbindlist(lapply(X = LibrarySizes, FUN = function(libsz){
  listi = replicate(N, simulate_community(OTUproportions, libsz), simplify=FALSE)
  dti = rbindlist(listi)
  dti[, LibrarySize := as.character(libsz)]
  return(dti)
}))
```


&nbsp;

---

# Graphic Summaries

## Scatter plot of individual results

```{r}
title = "Accuracy of Individual Simulations"
p + geom_point(aes(color=LibrarySize, fill=LibrarySize), data=bigdt,
               position=position_jitterdodge(), size=1) + 
  ggtitle(title)
title = "Accuracy of Aggregate Simulations (boxplot)"
p + geom_boxplot(aes(fill=LibrarySize),
                 data=bigdt, 
                 position="dodge") +
  ggtitle(title)
```

&nbsp;

## Residuals v. Counts

```{r}
bigdt[, TrueProportion := spdf[Taxa, "proportion"]]
bigdt[, Residual := proportion - TrueProportion]
bigdt[, Error := abs(Residual)]
bigdt[, StdErr := Residual / proportion]
```

```{r}
ggplot(bigdt, aes(x = N,
                  y = Residual,
                  color = factor(TrueProportion),
                  shape = LibrarySize)) + 
  geom_point(size = 2, alpha = 0.5) + 
  scale_x_log10() + 
  annotation_logticks(sides = "b") +
  ggtitle("Residual v. Absolute Count")
```


&nbsp;

## Variance v. Counts

Graphically summarize variance, standard deviation, and error.

First prepare data for plotting

```{r}
# Summarize by trial and Taxa
bigdt2 = ddply(bigdt, c("LibrarySize", "Taxa"), function(x){
  data.frame(Variance = var(x$N),
             standardDeviation = sd(x$N),
             meanCount = mean(x$N),
             meanProp = mean(x$proportion),
             L = as.integer(x$LibrarySize[1])
             )
})
bigdt2 <- data.table(bigdt2)
bigdt2[, TrueProportion := spdf[Taxa, "proportion"]]
bigdt2[, cv := Variance / as.numeric(LibrarySize)]
```

Define plots

```{r}
# Plots
title = "Accuracy and Library Size"
p + geom_boxplot(aes(fill=LibrarySize), data=bigdt, position="dodge") +
  geom_errorbar(data=bigdt2, position="dodge",
                aes(x=Taxa, y=NULL, colour=LibrarySize,
                    ymax=meanProp+standardDeviation/L,
                    ymin=meanProp-standardDeviation/L)) + 
  ggtitle(title)

title = "Observed Variance versus Observed Mean Counts"
ggplot(data = bigdt2, 
       mapping = aes(x = meanCount,
                     y = Variance,
                     colour=LibrarySize)) + 
  geom_point(size=5, alpha=0.5) + 
  geom_abline(intercept = 0, slope = 1) + 
  scale_y_log10() + 
  scale_x_log10() + 
  annotation_logticks(sides = "lb") +
  ggtitle(title)

title = "Observed Variance versus Expected Variance (Poisson)"
ggplot(bigdt2, aes(L*TrueProportion, Variance, color=LibrarySize)) + geom_point(size=5, alpha=0.5) + 
  geom_abline(intercept = 0, slope = 1) + 
  scale_y_log10() + 
  scale_x_log10() + 
  annotation_logticks(sides = "lb") +
  xlab("Expected Variance [counts^2]") + 
  ylab("Estimated Variance [counts^2]") +
  ggtitle(title)
```

&nbsp;

## Error in Proportion Estimate v. Counts

Notice how proportion on its own 
is a very bad predictor of the uncertainty/error!

```{r}
title = "Weak error dependency on proportion value"
ggplot(data = bigdt2, 
       mapping = aes(x = TrueProportion,
                     y = abs(TrueProportion - meanCount/L) / TrueProportion,
                     color = LibrarySize)) + 
  geom_point(size=5, alpha=0.5) + 
  scale_y_log10() + 
  scale_x_log10() + 
  annotation_logticks(sides = "lb") +
  xlab("Observed Proportion") + 
  ylab("Error [Proportion]") +
  ggtitle(title)
```


```{r}
title = "Weak error dependency on proportion"
ggplot(data = bigdt, 
       mapping = aes(x = proportion, 
                     y = log2(proportion / TrueProportion), 
                     colour = LibrarySize)) + 
  geom_hline(yintercept = 0, size = 0.25) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  ylab("Log2(Observed / Truth)") +
  annotation_logticks(sides = "b") +
  ggtitle(title)
```


Meanwhile the uncertainty is actually 
highly structured with respect to counts
(it is Poisson after all).

```{r}
title = "Error strongly dependent on count"
ggplot(data = bigdt, 
       # mapping = aes(N, StdErr, colour = LibrarySize)) + 
       mapping = aes(N, log2(proportion / TrueProportion), colour = LibrarySize)) + 
  geom_hline(yintercept = 0, size = 0.25) +
  geom_point(alpha = 0.5) +
  ylab("Log2(Observed / Truth)") +
  xlab("Count") +
  scale_x_log10() +
  annotation_logticks(sides = "b") +
  ggtitle(title)
```

```{r}
title = "Error strongly dependent on count"
ggplot(data = bigdt2, 
       mapping = aes(x = meanCount,
                     y = abs(TrueProportion - meanCount/L) / TrueProportion,
                     color=LibrarySize)) + 
  geom_point(size=5, alpha=0.5) + 
  scale_y_log10() + 
  scale_x_log10() + 
  annotation_logticks(sides = "lb") +
  xlab("Mean Count") + 
  ylab("Error [Proportion]") +
  ggtitle(title)
```


```{r}
title = "Variance in Proportion Estimate v. Observed Mean Counts"
ggplot(data = bigdt[, .(meanCount = mean(N),
                        VarProp = var(proportion)), 
                    by = .(LibrarySize, Taxa)], 
       mapping = aes(meanCount, VarProp, colour=LibrarySize)) + 
  geom_point(size=5, alpha=0.5) + 
  # geom_abline(intercept = 0, slope = 1) + 
  scale_y_log10() + 
  scale_x_log10() + 
  annotation_logticks(sides = "lb") +
  ggtitle(title)
```


---

&nbsp;


# Conclusion

The uncertainty due to sampling is actually 
highly structured with respect to counts,
following the Poisson distribution.
For intuition, real experiments have shown
that technical replicates
(different sequencing runs of the same prepared sample)
are also well described by the Poisson.
It is clear from this simulation that information 
represented by the count of each taxa in each sample
is highly predictive of the uncertainty 
in the corresponding proportion.
Converting to proportions early 
and ignoring the discrete counts is perilous,
especially at low values (e.g. rare taxa).

