---
title: "UW Summer Institutes - Introduction to Metagenomic Data Analysis - Lab 5"
output: html_notebook
---

# Setup

```{r load_packages}
library(phyloseq)
library(vegan)
library(ade4)

library(ggplot2)
library(ggthemes)
library(stringr)
library(tidyr)
library(dplyr)
```

\  

# Load the provided RData image of Nature 488, pp. 621-626 data

```{r load_data}
load("../data/STAT.rdata")
```

\  

# Compute Jensen-Shannon Divergence distance of the fecal samples.

```{r fecal_phy}
fecal_phy <- subset_samples(phy, Location == 'fecal')
fecal_pheno <- phenotypes

rownames(fecal_pheno) = paste("fecal", rownames(phenotypes), sep = "_")
fecal_phy = merge_phyloseq(fecal_phy, 
                            sample_data(fecal_pheno))
```


```{r dist_jsd}
dist_jsd <- phyloseq::distance(fecal_phy, method = distanceMethodList$JSD)
str(dist_jsd)
```

\  

# Use `meandist()` to calculate intra- and inter- group distance for the Treatment factor variable. Plot the barplot of intra-group distances.

```{r treat_dist}
treat_dist <- with(sample_data(phy), meandist(dist_jsd, Treatment))
treat_dist %>% 
    diag() %>% 
    as.data.frame() %>% 
    rename_(intragroup_dist = ".") %>% 
    add_rownames("treatment") %>% 
    ggplot(aes(x = treatment, y = intragroup_dist)) +
    geom_bar(stat = "identity")
```

\  

# Use `dudi.pco()` to perform PCoA of the JSD distances of fecal samples

```{r}
pco <- dudi.pco(cailliez(dist_jsd), 
                scannf = FALSE, nf = 2)
names(pco)
screeplot(pco)
```

## How many components do we want to keep? Why?

\  

# Calculate percent inertia explained by each component kept.

```{r calc_perc_inertia}
pco$eig[1:2]/sum(pco$eig)
```


\  

# Use `s.class()` function to plot the projection of the data with Treatment variable.

```{r plot_proj_treatment}
s.class(pco$li, sample_data(fecal_phy)$Treatment)
```


\  

# Use `s.value()` function to plot additional phenotypes on the projection of the data.

```{r plot_proj_gip}
s.value(pco$li, sample_data(fecal_phy)$GIP, sub = "GIP")
```

```{r plot_proj_insulin}
s.value(pco$li, sample_data(fecal_phy)$Insulin, sub = "Insulin")
```

\  

# Using `cor` function explore correlations of the additional phenotypes with the 
principal coordinates.

```{r cor_mat}
cor_mat = with(sample_data(fecal_phy),
               cor(cbind(PCo = pco$li,
                         GIP, Insulin, Leptin,
                         IGF1, BMD, mFat, pFat)))

round(cor_mat[, 1:2], digits = 2)
```

\  

# Plot correlations in #11.

```{r plot_cor_mat}
library(viridis)
cor_mat[, 1:2] %>% 
    as.data.frame() %>% 
    add_rownames("correlate") %>% 
    gather(pc, value, `PCo.A1`, `PCo.A2`) %>% 
    ggplot(aes(x = correlate, y = pc)) +
    geom_tile(aes(fill = value)) +
    geom_text(aes(label = round(value, digits = 2))) +
    scale_fill_viridis()
```


\  

# Perform PCoA analysis on the entire dataset (not just fecal samples) and produce 
plots with respect to Location and Treatment variables and their interactions.

\  

# How do we remove the effect of the Location variable in PCoA analysis? TIP: `?wca`